The Key to a Data Parallel Compiler

### 2016-04-11 2016-spring-array/2016-array-key.tex
Incorporate David's advice and some spelling corrections

We present a language-driven strategy for creating an inherently parallel 
compiler. We select a set of well known data-parallel primitives, and then 
restrict program construction to the composition of these operations into 
functions, without any forms of non-linear control flow, such as branching, 
recursion, or pattern matching. Programs constructed in such a style are 
data-parallel by construction. A key problem to the construction of compilers 
in such a restricted environment is dealing with the inherently recursive and 
nested structure of the AST. By working over a linearized representation of 
the AST as a matrix, combined with a structure we call a Node Coordinate 
Matrix, and using data-parallel primitives, we are able to tame the recursive 
and nested nature of the AST into something that can be processed handily 
using only data-parallel, data-flow programs, without requiring branching 
or other forms of non-data parallel control flow.

@@

\item We situation this technique into a broader, language-driven 
strategy for compiler construction that enables the development of parallel 
by construction compilers that are high-level in their implementation and 
exceptionally concise.

@@

The Key operater (written \verb;⌸;) is central to our strategy for compilation.
It allows us to use the Node Coordinates to their full effect. 
The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding 
keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in 
\verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;. 
The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
A simple example of this at work is to compute a histogram using Tally (\verb;≢;):

@@

We make a few notes on the table and the analysis. The Key operator
can be used with or without the node coordinates structure. In cases
where the AST is sufficiently flat and arranged in a sufficiently
convenient order, the grouping operations are very simple. In these
cases, the Key operator amounts to a map operation over some selected
sub-trees selected at a specific depth. These occur later in the
compiler where the reader will note appearences of the Key operator
without the use of Node Coordinates.

@@

The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the Key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (⍤) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community \cite{bernecky1987rank,hui1995rank}.

Fritz Henglein demonstrated a class of operations, called
discriminators, of which the Key operator is a member.  
\cite{henglein2013dd}
Namely,
a discriminator performs the same grouping computation as Key, but
does not apply a function over these groups with their keys. Henglein
provides a linear implementation of these operations.

@@

We have derived a method of performing computation over sub-trees
selected on the basis of inter-node properties through the use of the
Key (⌸) operator and node coordinates, which enable local computation
of these inter-node properties. This method is both general and direct,
and when combined with traditional and more mundane array programming,
suffices to implement the complete core of a compiler, modulo parsing
and code generation. The method requires no special operations or
unique special casing primitives in the language. Moreover, it is
strictly data-parallel and data-flow, without any complex control flow,
which results in exceptionally concise code. These methods permit a 
general, high-level approach to constructing a compiler that is 
parallel by construction by constricting the language to only a data parallel 
subset of a normal array language.

We have demonstrated the technique and the core insights behind the
data structures involved. It presents a solution to a very old and
traditional problem in a very uncommon light, by eschewing the common
practices that underlie every other significant and general solution
found in modern compilers today and replacing them with an entirely
different paradigm centred on parallelism and aggregate operations.

>>>>

We present a language-driven strategy for creating an inherently parallel 
compiler. We select a set of well known data-parallel primitives, and then 
restrict program construction to the composition of these operations into 
functions, without any forms of non-linear control flow, such as branching, 
recursion, or pattern matching. Programs constructed in such a style are 
data-parallel by construction. A key problem to the construction of compilers 
in such a restricted environment is dealing with the inherently recursive and 
nested structure of the AST. By working over a linearized representation of 
the AST as a matrix, combined with a structure we call a Node Coordinate 
Matrix, and using data-parallel primitives, we are able to tame the recursive 
and nested nature of the AST into something that can be processed handily 
using only data-parallel, data-flow programs, without requiring branching 
or other forms of non-data parallel control flow. 
The methods are general and independent of the language being compiled.

@@

\item We situation this technique into a broader, language-driven 
strategy for compiler construction that enables the development of parallel 
by construction compilers that are high-level in their implementation, 
exceptionally concise, and independent of the language being compiled.

@@

First introduced in the J language \cite{hui2014key},
the Key operator (written \verb;⌸;) is central to our strategy for compilation.
It allows us to use the Node Coordinates to their full effect. 
The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding 
keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in 
\verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;. 
The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
Let's compute a histogram using Tally (\verb;≢;) and Key (\verb;⌸;):

@@

We make a few notes on the table and the analysis. The Key operator
can be used with or without the node coordinates structure. In cases
where the AST is sufficiently flat and arranged in a sufficiently
convenient order, the grouping operations are very simple. In these
cases, the Key operator amounts to a map operation over some selected
sub-trees selected at a specific depth. These occur later in the
compiler where the reader will note appearances of the Key operator
without the use of Node Coordinates.

@@

The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the Key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (\verb;⍤;) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community \cite{bernecky1987rank,hui1995rank}.

Fritz Henglein demonstrated a class of operations, called
discriminators, of which the Key operator is a member  
\cite{henglein2013dd}
, namely,
a discriminator performs the same grouping computation as Key, but
does not apply a function over these groups with their keys. Henglein
provides a linear implementation of these operations.

@@

We have derived a method of performing computation over sub-trees
selected on the basis of inter-node properties through the use of the
Key (\verb;⌸;) operator and node coordinates, which enable local computation
of these inter-node properties. This method is both general and direct,
and when combined with traditional and more mundane array programming,
suffices to implement the complete core of a compiler, modulo parsing
and code generation. The method requires no special operations or
unique special casing primitives in the language. Moreover, it is
strictly data-parallel and data-flow, without any complex control flow,
which results in exceptionally concise code. These methods permit a 
general, high-level approach to constructing a compiler that is 
parallel by construction by constricting the language to only a data parallel 
subset of a normal array language.

We have demonstrated the technique and the core insights behind the
data structures involved. It presents a solution to a very old and
traditional problem in a very uncommon light, by eschewing the common
practices that underlie every other significant and general solution
found in modern compilers today and replacing them with an entirely
different paradigm centered on parallelism and aggregate operations.
